\plaintitle

### 2020-03-22 wip-design/paper.tex
Adding implementation section

@@

\subsection{The Gamma Language}
In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
or an expression that evaluates to a value such as a data table or a chart.
An expression is a reference to a data source followed by a chain of member accesses.
A member can be either an ordinary member or an operation which takes a list of parameters
enclosed in parentheses. Our implementation supports a few other features, but those
are not used in this paper.

The language uses a type system to infer what members are available at a given point in a chain
of members. The type is an object with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by a type provider for a particular data source.
The types are also generated lazily. A type is only computed when the corresponding member
is actually used in code.

@@

\subsection{Data Cube Type Provider}

@@

\subsection{Graph Database Type Provider}

\newpage

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/sidebyside}
\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
of China and United States.}
\label{fig:sidebyside}
\vspace{-0.5em}
\end{figure}

>>>>

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/sidebyside}
\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
of China and United States.}
\label{fig:sidebyside}
\end{figure}

@@

\subsection{The Gamma Language}
In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
or an expression that evaluates to a value such as a data table or a chart.
An expression is a reference to a data source followed by a chain of member accesses.
A member can be either an ordinary member or an operation which takes a list of parameters
enclosed in parentheses. When the member name contains non-alphanumerical characters such as
whitespcae, it needs to be written in quotes. Our implementation supports a few other features,
but those are not used in this paper.

The Gamma uses a type system to infer what members are available at a given point in a chain.
A type is an object with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by a type providers for individual data sources.
The types are generated lazily and are only computed when a member appears in code.

@@

\begin{figure}
\centering
\includegraphics[scale=0.28]{figures/cubetp}
\caption{Exploring data cubes. The World Bank provider allows users to first choose
  a country or a year and then an indicator.}
\label{fig:cubetp}
\end{figure}

\subsection{Data Cube Type Provider}
Our first type provider allows users explore data form a data cube. For example, the World
Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
countries in the world each year. The data set is a three-dimensional cube with dimensions
corresponding to countries, indicators and years. The following example, also shown in
Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:

\begin{thegamma}
worldbank.byCountry.'United States'
  .'Climate Change'.'CO2 emissions (kt)'
\end{thegamma}

As illustrated in Figure~\ref{fig:cubetp}, the provider allows uses to select a data series
from the data cube. By choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
we restrict the cube to a two-dimensional plane. We then choose an indicator category
\ikvd{\textquotesingle Climate Change\textquotesingle} and a specific indicator
\ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}, obtaining a time series with
years as keys and emission data as values. The World Bank provider also supports filtering
by year and indicator.

Another type provider available in The Gamma that is based on the data cube structure allows
the user to explore UK government expenditure:

\begin{thegamma}
expenditure.byService.Defence.inTermsOf.GDP
\end{thegamma}

The dimensions of the cube are government services, years and value type (adjusted, nominal,
per GDP). Here, we select the \ikvd{Defence} service and \ikvd{GDP} value type. Our implementation
does not yet use a standardized data cube storage and so adding another data cube source
currently requires implementing a new type provider.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. It is based on the theory developed Petricek~\cite{dotdriven}. Unlike the
data cube provider, the provider for tabular data does not just allow selecting a subset of the
data, but it can be used to construct SQL-like query. Consider the example from Figure~\ref{fig:thegamma}:

\begin{thegamma}
olympics.'filter data'.'Games is'.'Rio (2016)'.then
  .'group data'.'by Team'.'sum Gold'.'sum Silver'.then
  .'sort data'.'by Gold descending'
\end{thegamma}

The example works with a CSV file that records individual medals awarded in Olympic games.
The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
calculates total number of gold and silver medals for each team (country) before sorting the data.

When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
determine the operation type. Those are followed by operation parameters. For example, when grouping
data, we first select the key and then add a number of aggregations to calculate over the group.
Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
calculating the sum, average or the number of distinct values. As illustrated in
Section~\ref{sec:cases}, this still allows us to construct a wide range of practical queries.

\begin{figure}
\centering
\includegraphics[scale=0.28]{figures/graphtp}
\vspace{0.5em}
\caption{Exploring graph databases. The user specifies a path through the data, possibly with
  placeholders to select multiple nodes.}
\label{fig:graphtp}
\end{figure}

\subsection{Graph Database Type Provider}
Our third type provider allows users explore data from graph databases. A graph database
consists of nodes representing entities and relationships between them. In the context of
data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.

The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

\begin{thegamma}
drwho.Character.Doctor
  .'ENEMY OF'.'[any]'
  .'APPEARED IN'.'Day of the Moon'
\end{thegamma}

We start from the \ikvd{Doctor} node and then follow two relationships. We use
\ikvd{\textquotesingle ENEMY OF\textquotesingle.\textquotesingle [any]\textquotesingle}
to follow links to all enemies of the Doctor and then specify
\ikvd{\textquotesingle APPEARED IN\textquotesingle.\textquotesingle Day of the Moon\textquotesingle}
to only select only such enemies that appear in a specific episode. The resulting query
is illustrated in Figure~\ref{fig:graphtp}.

The provider works with any graph database and generates members automatically, based on the
data in the database. In the above example, \ikvd{ENEMY OF} and \ikvd{APPEARED IN} are labels
of relations and \ikvd{Doctor} and \ikvd{Day of the Moon} are labels of nodes. The
\ikvd{[any]} member defines a placeholder that can be filled with any node with the specified
relationships. The results returned by the provider is a table of properties of all nodes
along the specified path. As discussed in the next section, such table can be further queried
using the tabular data type provider.
